---
title: "Introduction to MSanalyzeNOM"
author: Robert B. Young
date: "5/30/2020"
output: html_notebook
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set global options, include = FALSE}
options(digits = 8)
```

```{r load packages, include = FALSE}
library(MSanalyzeNOM)
```

```{r set sample data file location}
DOM_xcel <- system.file("extdata", "DOM_Sample.xls", package = "MSanalyzeNOM", mustWork = TRUE)
```



## Introduction

This document demonstrates R functions in the ***{MSanalyzeNOM}*** package for analyzing the mass spectrometry (MS) data of natural organic matter (NOM) samples after formula assignment. It was specifically created to analyze MS data from NOM samples analyzed by Fourier transform ion cyclotron resonance (FTICR) MS at the [National High Magnetic Field Laboratory](https://nationalmaglab.org/user-facilities/icr) (MagLab) in Tallahassee, Florida, USA. Usually, the analyses were  conducted using electrospray ionization in negative ion mode (NegESI), which is the default ion type in the {MSanalyzeNOM} package. Additional features will be added to deal with electrospray ionization in positive ion mode (PosESI), and odd electron ions like those produced during atmospheric pressure photoionization (APPI). In addition, the {MSanalyzeNOM} package generally assumes that the detected ions are singly-charged, based on literature describing the ESI-FTICR MS analysis of dissolved NOM (e.g., Stenson et al. 2002). 



## Reading MS Data from the MagLab

The {MSanalyzeNOM} package uses the ***get_sample_data()*** function to read Excel files exported by PetroOrg&copy; software, which was created at the MagLab for mass calibration, molecular formula (elemental composition) assignment, and other purposes. The Excel file contains one or more sheets with summary data, a sheet with mass spectrometry data for detected ions that could not be assigned molecular formulas ("no hits"), and numerous sheets containing mass spectrometry data for the assigned molecular formulas, which have been grouped by heteroatom class (the number and identity of heteroatoms contained in the assigned molecular formulas). For example, C<sub>6</sub>H<sub>10</sub>O<sub>5</sub>, C<sub>7</sub>H<sub>12</sub>O<sub>5</sub>, and C<sub>8</sub>H<sub>14</sub>O<sub>5</sub> would be part of the "O5" heteroatom class, and C<sub>11</sub>H<sub>9</sub>NO<sub>6</sub>S would be part of the "N1 O6 S1" heteroatom class.

The ***get_sample_data()*** function returns a named list containing file and sample information, selected mass spectrometry parameters, mass spectrometry data related to the assigned molecular formulas, and mass spectrometry data related to detected ions that could not be assigned molecular formulas. Most of other functions in the {MSanalyzeNOM} package rely on column names produced by the ***get_sample_data()*** function. The following example illustrates the data that is provided for the assigned molecular formulas: 

```{r get_sample_data, message = FALSE}
DOM_sample <- get_sample_data(DOM_xcel)
DOM_sample$sample_name
DOM_sample$ion_technique
DOM_sample$element_list
```

```{r results_get_sample_data, message = FALSE}
head(DOM_sample$assigned_formulas)
```

The default list of elements includes C, H, N, O and S, but a different element list can be specified if it was used for formula assignment. 

```{r message = FALSE}
sample_data <- get_sample_data(DOM_xcel, element_list = c("C", "H", "N", "O"))
sample_data[["element_list"]]
```

Additional details can be examined by typing *"?get_sample_data"* in the R console.

```{r, include = FALSE}
rm(sample_data)
```



## Visualizing the Data from a Single Sample

Major biogeochemical classes of chemical compounds, such as lipids and carbohydrates, have characteristic H/C and O/C ratios (Kim et al. 2003). As a result, van Krevelen diagrams, which typically plot H/C vs. O/C ratios, have become an important tool for characterizing NOM. 


### 1. Calculating the Elemental Ratios

The {MSanalyzeNOM} package uses the ***get_elemental_ratios()*** function to compute H/C and O/C ratios. The desired numerator and denominator are passed to the function to create a new column with the desired ratio (e.g., "HtoC"). "C" is the default denominator for ***get_elemental_ratios()***.

```{r get_elemental_ratios}
DOM_sample$assigned_formulas <- DOM_sample$assigned_formulas %>%
  get_elemental_ratios(num = "H") %>%
  get_elemental_ratios(num = "O")
```

```{r results_get_elemental_ratios, echo = FALSE}
DOM_sample$assigned_formulas %>%
  dplyr::select(-hetero_class, 
                -(mz:S)) %>%
  head()
```



### 2. Creating a van Krevelen Diagram

A traditional van Krevelen diagram plots O/C vs. H/C. In general, high H/C ratios reflect a high degree of saturation, and lower H/C ratios reflect more rings and double bonds. Similarly, O/C ratios reflect the relative presence of hydroxyl, carboxyl and other oxygen functional groups. Accordingly, the area above the horizontal line at H/C = 1.5 has been designated as the aliphatic region in accordance with D'Andrilli et al. 2015 and Lv et al. 2017, and the area below the diagonal line from H/C = 1.1 has been  designated as the aromatic region, based on a modified aromaticity index \geq 0.5 in accordance with Koch & Dittmar 2006.

The following plot, created by the ***make_VK_flat()*** function, shows the distribution of the assigned formulas as a function of their elemental ratios, but contains no information about the intensities of the detected ions. The darker areas are where the highest number of formulas are plotted.

```{r make_VK_flat, fig.height = 4.5, fig.width = 5.5}
plot_VK_flat(DOM_sample$assigned_formulas, plot_title = DOM_sample$sample_name)
```

The numbers of detected ions and assigned formulas can be influenced by ion suppression, making van Krevelen diagrams from different samples or analyses difficult to compare. In fact, the vast majority of detected ions and assigned formulas occur at relatively low abundances, and are sensitive to ion suppression. In addition, because the vast majority of detected ions and assigned formulas occur at relatively low abundances, a color scheme based on a gradient of the assigned formulas' relative abundances doesn't necessarily improve resolution.

```{r make_VK_grad, fig.height = 4.5, fig.width = 6.5}
plot_VK_grad(DOM_sample$assigned_formulas, plot_title = DOM_sample$sample_name)
```

One alternative for increasing resolution is to group assigned formulas by their percent contribution to total assigned abundance, and then plot the grouped formulas in a van Krevelen diagram. For example, the formulas can be ordered from least to most abundant. Then,  the formulas that account for the first 25% of total abundance can be grouped together (the "bottom 25%"), the formulas that account for the next 25% of total abundance can be grouped together, and so on, until four distinct groups have been formed: "top 25%", "second 25%", "third 25%", and "bottom 25%". In the current example, the vast majority of assigned formulas comprise the bottom 25%, and a comparatively small number of formulas comprise the top 25%. The end result is that the van Krevelen diagram better shows where the most abundant assigned formulas congregate.

[create functions for adding perc_abundances, adding cum_perc_abundances and grouping, and arranging and plotting VK diagrams colored by group (make_VK_group_25perc)]

```{r get_25perc_groups}
DOM_sample$assigned_formulas <- DOM_sample$assigned_formulas %>%
  get_perc_abund() %>%
  get_25perc_groups()
```

```{r results_get_25perc_groups, message = FALSE}
DOM_sample$assigned_formulas  %>% 
  dplyr::group_by(`25perc_group`) %>% 
  dplyr::summarize(num_formulas = dplyr::n())
```

```{r, fig.height = 4.5, fig.width = 6.5}
plot_VK_25perc_groups(DOM_sample$assigned_formulas, plot_title = DOM_sample$sample_name)
```

